---
title: "Visualizing Data with the Heatmap Function"
author: "Yichao Hua"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing Data with the Heatmap Function}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Introduction

The `Heatmap` function provides a flexible and comprehensive way to visualize matrices, especially those produced by the `CalcStats` function. This vignette provides a quick overview of how to utilize the various features and capabilities of the `Heatmap` function to generate customized visualizations.

## Basic Usage

First, let's generate a sample matrix using the `CalcStats` function:

```{r, message=FALSE, warning=FALSE}
library(Seurat)
library(SeuratExtend)

# Assuming pbmc data and VariableFeatures function are available
genes <- VariableFeatures(pbmc)
toplot <- CalcStats(pbmc, features = genes, method = "zscore", order = "p", n = 5)
```

Now, we can produce a basic heatmap:

```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=7}
Heatmap(toplot, lab_fill = "zscore")
```

## Customizing the Heatmap

### Adjusting Color Schemes

The `color_scheme` parameter allows for flexibility in visualizing data. Here are some ways to change the color theme of your heatmap:

```{r, fig.width=5, fig.height=7}
# White to dark green
Heatmap(toplot, lab_fill = "zscore", color_scheme = c("white", muted("green")))

# Dark blue to light yellow (centered at 0) to dark red
Heatmap(toplot, lab_fill = "zscore", color_scheme = c(
  low = muted("blue"),
  mid = "lightyellow",
  high = muted("red"))
)
```

You can also use predefined color schemes, such as those from the viridis package:

```{r, fig.width=5, fig.height=7}
Heatmap(toplot, lab_fill = "zscore", color_scheme = "A")
```

### Modifying Axis and Labels

If the first name on the x-axis is too long and surpasses the left boundary of the plot, you can expand the x-axis's left boundary:

```{r, fig.width=5, fig.height=7}
Heatmap(toplot, lab_fill = "zscore", expand_limits_x = -0.5)
```

For denser matrices, you may wish to only show a subset of gene names:

```{r, fig.width=5, fig.height=7}
toplot2 <- CalcStats(pbmc, features = genes[1:500], method = "zscore", order = "p")
Heatmap(toplot2, lab_fill = "zscore", feature_text_subset = genes[1:20], expand_limits_x = c(-0.5, 11))
```

### Faceting the Heatmap

You can also split the heatmap based on gene groups:

```{r, fig.width=5, fig.height=7}
gene_groups <- sample(c("group1", "group2", "group3"), nrow(toplot2), replace = TRUE)
Heatmap(toplot2, lab_fill = "zscore", facet_row = gene_groups) +
  theme(axis.text.y = element_blank())
```

