---
title: "What's New in v1.1.0"
author: "Yichao Hua"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What's New in v1.1.0}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## New Features and Enhancements in v1.1.0

### New Visualization Functions

#### Enhanced Dot Plots with `DotPlot2`

A new function `DotPlot2` has been introduced, combining dot size (percent of expressing cells) with color intensity (average expression) for more informative visualizations:

```{r, message=FALSE, warning=FALSE, fig.width=6, fig.height=4.5}
library(Seurat)
library(SeuratExtend)

# With grouped features
grouped_features <- list(
  "B_cell_markers" = c("MS4A1", "CD79A"),
  "T_cell_markers" = c("CD3D", "CD8A", "IL7R"),
  "Myeloid_markers" = c("CD14", "FCGR3A", "S100A8")
)
DotPlot2(pbmc, features = grouped_features)
```

#### New Volcano Plots

The new `VolcanoPlot` function provides statistical visualization of differential expression:

```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
VolcanoPlot(pbmc, 
            ident.1 = "B cell",
            ident.2 = "CD8 T cell")
```

#### UMAP Arrow Annotations

Added `theme_umap_arrows` for simplified axis indicators on dimension reduction plots:

```{r, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
DimPlot2(pbmc, theme = NoAxes()) + theme_umap_arrows()
```

### Enhanced Existing Functions

#### VlnPlot2 Enhancements

Added support for mean and median lines in violin plots:

```{r, message=FALSE, warning=FALSE, fig.width=6, fig.height=6.5}
VlnPlot2(pbmc, 
         features = c("CCR7", "IL7R", "TCF7"),
         cols = "light",
         show.mean = TRUE,      # Show mean and median lines
         mean_colors = c("red", "blue"),  # Colors for mean and median
         ncol = 1)
```

### New WaterfallPlot Style: "segment"

Added a new visualization style "segment" to WaterfallPlot, providing an alternative way to display differences between conditions:

```{r, message=FALSE, warning=FALSE, fig.width=7, fig.height=7.2}
# Prepare data
pbmc <- GeneSetAnalysis(pbmc, genesets = hall50$human)
matr <- pbmc@misc$AUCell$genesets

# Create a plot using the new segment style
WaterfallPlot(
  matr,
  f = pbmc$cluster,
  ident.1 = "Mono CD14",
  ident.2 = "CD8 T cell",
  style = "segment",
  color_theme = "D"
)
```

### Color Scheme Updates

#### New Default Color Schemes

Two major color scheme changes have been implemented in v1.1.0:

1. For continuous variables: Changed from viridis "A" to RColorBrewer "Blues"
2. For heatmaps: Updated from `c(low = muted("blue"), mid = "white", high = muted("red"))` to "BuRd"

Here are visual comparisons of the old and new defaults:

##### Continuous Variable Color Scheme

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3.5}
# Create a side-by-side comparison for continuous variables
library(cowplot)
library(ggpubr)

# Old default (viridis "A")
p1 <- DimPlot2(pbmc, 
               features = "CD3D", 
               cols = "A",  # Old default
               theme = theme_umap_arrows()) 

# New default (Blues)
p2 <- DimPlot2(pbmc, 
               features = "CD3D", 
               theme = theme_umap_arrows()) 

plot_grid(
  annotate_figure(p1, top = text_grob("Old Default (viridis 'A')", size = 14)), 
  annotate_figure(p2, top = text_grob("New Default ('Blues')", size = 14)),
  ncol = 2)
```

##### Heatmap Color Scheme

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
# Calculate data for heatmap
genes <- VariableFeatures(pbmc)
toplot <- CalcStats(pbmc, features = genes, method = "zscore", order = "p", n = 4)

# Create side-by-side heatmap comparison
p1 <- Heatmap(toplot, 
              color_scheme = c(low = scales::muted("blue"), 
                             mid = "white", 
                             high = scales::muted("red")), # Old default
              lab_fill = "zscore") +
      ggtitle("Old Default (blue-white-red)")

p2 <- Heatmap(toplot, 
              lab_fill = "zscore") +  # New default (BuRd) is automatically applied
      ggtitle("New Default ('BuRd')")

plot_grid(p1, p2, ncol = 2)
```

To revert to previous color schemes:
- For continuous variables: Use `cols = "A"` 
- For heatmaps: Use `color_scheme = c(low = scales::muted("blue"), mid = "white", high = scales::muted("red"))`

#### New RColorBrewer Integration

Added support for RColorBrewer sequential and diverging palettes:

- Sequential: "Blues", "BuGn", "BuPu", "GnBu", etc.
- Diverging: "BrBG", "PiYG", "PRGn", "PuOr", etc.
- Custom diverging: "GnYlRd", "BuYlRd", "GyRd", "BuRd", "PuOr"

### Feature Enhancements

- **VlnPlot2**: Now supports both `stats.method` and `stat.method` as parameter inputs ([#10](https://github.com/huayc09/SeuratExtend/issues/10))
- **ClusterDistrBar**: Added `reverse_order` parameter to adjust the stacking order ([#11](https://github.com/huayc09/SeuratExtend/issues/11))
- **WaterfallPlot**: Set upper limit for -log10(p) values to avoid NA issues ([#14](https://github.com/huayc09/SeuratExtend/issues/14))
- **DimPlot2**: Improved automatic point size adjustment and fixed point display issues in raster mode ([#17](https://github.com/huayc09/SeuratExtend/issues/17))
- **show_col2**: Function is now exported, allowing users to knit Visualization.Rmd without issues ([#8](https://github.com/huayc09/SeuratExtend/issues/8))

### Bug Fixes

- **VlnPlot2**: Now explicitly uses `dplyr::select` internally to avoid conflicts with other packages' select functions ([#5](https://github.com/huayc09/SeuratExtend/issues/5), [#10](https://github.com/huayc09/SeuratExtend/issues/10))
- **ClusterDistrBar**: Fixed display issues when factor levels are numeric (e.g., seurat_clusters)

### Documentation Updates

#### Conda Environment Setup

The `create_condaenv_seuratextend()` function documentation has been updated with important compatibility information:

- Currently supported and tested on:
  * Windows
  * Intel-based macOS (not Apple Silicon/M1/M2)
  * Linux (Ubuntu 20.04)

**Note for Apple Silicon Users**: The function is not currently compatible with Apple Silicon/M1/M2 devices ([#7](https://github.com/huayc09/SeuratExtend/issues/7)). Users with Apple Silicon devices who are interested in contributing to the development of M1/M2 support are welcome to reach out via GitHub Issues.

#### Windows-Specific File Download

When downloading loom files (which are HDF5-based binary files) on Windows, it's essential to use `mode = "wb"` in the download.file() function:

```{r eval=FALSE}
# Example for Windows users
download.file("https://example.com/file.loom", "file.loom", mode = "wb")
```

This prevents Windows from modifying line endings in the binary file, which would corrupt the HDF5 format. Mac and Linux users don't require this parameter, but including it is harmless.

```{r}
sessionInfo()
```
